---
title: "R Syntax & Core Data Structures"
format: html
author: 
  - name: "Devon J. Boland"
    orcid: "0000-0001-8903-9718"
    url: "https://devonjboland.github.io/"
    affiliation: "Texas A&M Institute for Genome Sciences & Society (TIGSS) - Bioinformatics Core, Texas A&M University Vashist College of Medicine"
    email: "devonjboland@tamu.edu"
    corresponding: true
date: last-modified
code-annotations: below
---
[Download this Notebook (R-Introductions.qmd)](R-coding.qmd){.btn .btn-primary}

## R Basics
Congratulations, you have made it through the through the fundamentals of setting up your own personal R environment, 
and we have hardly done any coding. If you have managed to not fall asleep so far, get ready for the excitement of writing code! 
**Join the dark side of the lab, and feel it's power** (*read in a legally distinct and not trademarkable character's voice*)!

### Hello, world!
In all seriousness though, coding can be as simple or complex as you make it. For instance, we are going to begin by writing the first program everyone learns in any programming language, printing the `Hello, world!` statement. We will use the `print()` function from base R. We will go over each part of it shortly. Begin, typing the following in your R script:
```{r}
print("Hello, world!")
```
As you can see the statement "Hello, world!" was output (or printed, hence the function name) below in our R console. 
We called the `print()` function which takes as input a string (text), number, or even a variable as input. This `print()` function is used a lot across many programming languages. You have commonly encountered it in your day-to-day jobs. When coding it is often used to print a status update, error message, warning, etc. to the user as feedback so we aren't just starting at a static screen wondering if something has happened or we have messed up.

Now we are going to add a layer to our `print()` code from earlier. What if we wanted to store a message to ourselves, 
say remember to drink our coffee while it is still hot, and view it later. We achieve this by creating a `variable` and 
casting (storing) a value in it using the `<-` operator.
```{r}
note <- "Remember, to drink your coffee, while it is hot!"  # <1>
print("Here we are going to print our note below:")
print(note) # <2>
```

1. this line stores our message about our coffee into the variable "note".
2. then we print it to stdout.

Congrats! You have learned two basics in the R programming language, how to call a function `print`, 
and how to store a variable with the `<-` operator. Now we are going to move on to store data in R.

### Vector, the simplest way to store multiple data
Above we stored our first data, a single string into the variable `note`. But what if we wanted to store a list of numbers?
In R, a series of data is called a **vector** (in R we also have list objects, which behave differently but can look similar). 
We begin a vector using the `c()` and separating elements in the vector with a **comma**. Below, let's create a vector of numbers 1-5.
```{r}
num_vec <- c(1,2,3,4,5)
print(num_vec)
```

As you can see, we stored a list of the numbers 1-5 in the variable called `num_vec`. Unlike before, our variable now stored as numbers,
specifically as integers. You'll notice how we dropped the `""` marks as we intended to store numbers instead of strings.

However we can mix data types in vectors, lets see what happens when we try to mix numbers and strings:
```{r}
mix_vec <- c(1, "the", 2, "cow", 3, "moos")
print(mix_vec)
```

### DataFrames: R's Excel Table Data Object
Vectors are very useful and a core object type in R. Many other useful objects are built on top of vectors. However, the most common type
of data we deal with in R will be DataFrames/Tibbles. DataFrames (or the modern equivalent Tibbles, used in tidyverse) are the R equivalent of tables in Excel.

Tibbles are not included in base R, like Dataframes are, so here we need to attach the tidyverse package we installed earlier.Copy and run the following code to generate a dataframe:

```{r}
library(tidyverse)

df <- data.frame(name=c("A", "B"), value = c(5, 10)) # <1>
print(df)

tib <- tibble(name=c("C", "D"), value = c(15, 20)) # <2>
print(tib)
```

1. Here we use the base R function data.frame() function to create a dataframe object
2. Here we use the tidyverse tibble() function to create a tibble object.

Now we have two separate frames of data. Next we will show you how we can combine them. This is useful if we have multiple datasets we have loaded individually and now need to combine. Here we will use the `dplyr` function `bind_rows()`.
```{r}
combined_df <- bind_rows(df, tib)
```

#### Accessing Data inside Dataframes/Tibbles

Now let's take a look on how to access rows, columns, and specific values using base R's indexing methods. First let's 
see how we can view the row and column names of a dataframe:
```{r}
print(rownames(combined_df)) # <1>
print(colnames(combined_df)) # <2>
```

1. rownames() returns a vector of the row index names, or numbers if no names are present
2. colnames() returns a vector of the column names, or numbers if no names are present.

Above we can see that, the row indexes are numbers, and our column names are "name" and "value". We can use the `rownames()` and `colnames()` functions to replace the old labels with new ones if we wanted to as well. Next lets see how we print specific rows:
```{r}
print(combined_df[1,])
print(combined_df[3,])
```

Here the numbers represent coordinates where the x-value corresponds to row number, and the y-value corresponds to a column number within the dataframe. By leaving the second value empty (after the comma) we print a single row. The same would be true if we switched the numbers around, and we would print a column instead. To access specific data, we use both coordinate values:
```{r}
print(combined_df[1,2])
```

#### Importing Tabular/CSV Data

Now that we understand the structure of dataframes/tibbles in R, let's learn how to import
our CSV/TSV/Excel-formatted data files as dataframes into R. Here we are going to leverage `read.csv()` from base R. Later we will work on data we pull from [iLab](https://my.ilabsolutions.com/account/login), for now please downlaod a sample data set from [here](https://github.com/DevonJBoland/ABRF-SC3-R_Workshop/blob/main/data/fy26_costs.csv). The `read.csv()` has a few parameters. We can conveniently see a quick help guide on any function of interest by using the `?` before any function call try it first below:
```{r}
?read.csv()
csv_data <- read.csv("data/fy26_costs.csv") # <1>
print(csv_data)

csv_data$Costs # <2>
```

1. Here we only need to pass the path to our file, in some cases, you may need to specify which columns you would like to keep, and/or skip loading into memory.
2. Here is another way we can print entire columns of a dataframe, by accessing the specific column name of the dataframe with `$`.

Here we have an example of a simple monthly breakdown of costs and sales. What if we wanted to quickly calculate gross revenue per month? We can, quickly in fact, and it is as intuitive as using a calculator:
```{r}
csv_data$Revenue <- csv_data$Sales - csv_data$Costs # <1>
print(csv_data)
```

1. Here we create a new column with `$` and name it **Revenue**. We then subtract the vector `csv_data$Costs` from `csv_data$Sales`.

We have quickly calculated our total revenue for FY26. But this is something that can be readily done in excel with a similar equation and then a click and drag down. However what if we wanted some quick statistics for each of our columns:
```{r}
summary(csv_data)
```

The `summary()` function take a dataframe or vector as input and quickly output basic statistical analysis of the distribution of the data. You'll notice that our Month column is a column filled with strings. Instead we get an overview of the length, number of unique identities, number of blank entries, and the minimum and maximum number of characters in a row. All other columns contain numerical values and so we get quick distribution summaries of our data.

As a final example of what we can do with R combining everything we have gone over in this notebook. Lets print a message telling us the sum total of our costs, sales, and revenue numbers for FY26:
```{r}
print(paste0("Total Costs: $", sum(csv_data$Costs)))
print(paste0("Total Sales: $", sum(csv_data$Sales)))
print(paste0("Total Revenue: $", sum(csv_data$Revenue)))
```

Now that we have gotten some exposure on loading, manipulating, and messing around with data in R. Let's move on to a real-world example in the next section!
