---
title: "Core Admin - Wrangling & Visualizing iLab Revenue"
format: html
author: 
  - name: "Devon J. Boland"
    orcid: "0000-0001-8903-9718"
    url: "https://devonjboland.github.io/"
    affiliation: "Texas A&M Institute for Genome Sciences & Society (TIGSS) - Bioinformatics Core, Texas A&M University Vashist College of Medicine"
    email: "devonjboland@tamu.edu"
    corresponding: true
date: last-modified
code-annotations: below
---
[Download this Notebook (R-ilab.qmd)](R-ilab.qmd){.btn .btn-primary}

## Session Overview
- 1:20 - 1:40 | Data Wrangling with `dplyr`
- 1:40 - 2:10 | Visualizaing Revenue Trands with `ggplot2` & Exploring Plots

## Setup & Required Packages
Before starting, load the required core packages from `tidyverse`
```{r}
library(tidyverse)
library(lubridate)
library(ggplot2)
```

## Data Wrangling with `dplyr`
In this segment, we process raw billing data from iLab systems. We will clean test accounts, focus on completed/paid transactions, use the standard R native pipe `|>`, and parse timestamp strings into useful data features (`Months` and `Quarter`).

### Downloading iLab Data
We are going to download our revenue reports from iLab for FY26. If you do not have access to this information in iLab, please use the data provided [here](INSERTe GITHUB LINK to DATA Folder).

### Load in iLab Data
Now let's load in our data as have done previously:
```{r}
ilab_raw <- read.csv("data/module3_FY26_full.csv", header = TRUE)
```

### Clean iLab Data
We have our iLab data loaded into memory and stored in the vairable `ilab_raw`. However, if we inspect the data we will notice it is full of `n/a`, or just empty data points throughout the columns of the dataframe. We need to clean up the data and get it in a state that is useful for reporting an more informative for plotting.

We are going to accomplish this using the `dplyr` package bundled with `tidyverse`. Our goals are to:

* Filter out internal test accounts `(Customer Name != "Test Account")`, and unpaid/cancelled invoices `(payment_status == "Paid")`.  
* Parse timestamps into `Month` and `Quarter` using `lubridate` functions.  
* Format date categories cleanly for downstream visualization.  

```{r}
cleaned_revenue_data <- ilab_raw |>
  filter(
    Customer.Name != "Test Account",
    Billing.Status == "billing_initialized"
  ) |>
  mutate(
    date = mdy_hm(Billing.Date),
    month_num = month(date),
    month_label = month(date, label = TRUE, abbr = TRUE),
    quarter = factor(paste0("Q", quarter(date)))
  ) |>
  arrange(date)

head(cleaned_revenue_data)
```

## Visualizing Revenue Trends with `ggplot2`
Now that our data is clean, we will aggregate revenue trends and build report-ready figures.

### Monthly Revenue (Bar Chart)
To identify seasonal dips, we aggregate total revenue by month.
```{r}
monthly_revenue <- cleaned_revenue_data |> 
  group_by(month_label) |> 
  summarise(total_revenue = sum(Total.Price), .groups = "drop")
head(monthly_revenue)
```
```{r}
p1 <- ggplot(monthly_revenue, aes(x = month_label, y = total_revenue)) +
  geom_col(fill = "orange", color = "white", alpha = 0.9) +
  scale_y_continuous(labels = scales::dollar_format(prefix = "$")) +
  labs(
    title="Total Monthly Revenue (FY26)",
    subtitle="Analysis of seasonal dips in core facility billing",
    x="Month",
    y="Revenue (USD)"
  ) +
    theme_minimal(base_size = 12) +
  theme(
    plot.title=element_text(face = "bold", size = 14),
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank()
  )

  p1
```

### Quartely Revenue By Institution Type (Stacked Bar & Line Charts)
Next, we evaluate revenue per quarter broken down by Institution Type (Internal vs. External)
```{r}
quarterly_institution_revenue <- cleaned_revenue_data |> 
  mutate(Institution.Type = if_else(
    Customer.Institute == "Texas A&M University",
    "Internal",
    "External"
  )) |> 
    group_by(quarter, Institution.Type) |> 
    summarise(total_revenue = sum(Total.Price), .groups = "drop")
  
head(quarterly_institution_revenue)
```
```{r}
p2 <- ggplot(quarterly_institution_revenue, aes(x = quarter, y = total_revenue, fill = Institution.Type)) +
  geom_col(position = "stack", width = 0.6) +
  scale_fill_manual(values = c("Internal" = "blue", "External" = "orange")) +
  scale_y_continuous(labels = scales::dollar_format(prefix = "$")) +
  labs(
    title = "Quarterly Revenue Breakdown by Institution Type",
    subtitle = "Stacked var view of internal vs. external revenue",
    x = "Quarter",
    y = "Revenue (USD)",
    fill = "Institution Type"
  ) +
    theme_minimal(base_size = 12) +
    theme(
      plot.title = element_text(face="bold", size = 14),
      legend.position = "bottom"
    )
p2
```

### Line Chart Option
```{r}
p3 <- ggplot(quarterly_institution_revenue, aes(x = quarter, y = total_revenue, color = Institution.Type, group = Institution.Type)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 3) +
  scale_color_manual(values = c("Internal" = "#2b5c8f", "External" = "#d95f02")) +
  scale_y_continuous(labels = scales::dollar_format(prefix = "$")) +
  labs(
    title = "Quarterly Revenue Trajectory",
    subtitle = "Comparison of growth trends between Internal and External/Affiliate clients",
    x = "Quarter",
    y = "Revenue (USD)",
    color = "Institution Type"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    legend.position = "bottom"
  )

p3
```

### Exporint Report-Ready Plots
To export high-resolution figures for reports or internal sharing, we use `ggsave()`.
```{r}
dir.create("output_plots", showWarnings = FALSE)

ggsave(
  filename = "output_plots/monthly_revenue_trend.pdf",
  plot = p1,
  width = 8,
  height = 5,
  units = "in",
  dpi = 300
)

ggsave(
  filename = "output_plots/quarterly_revenue_institution_stacked.png",
  plot = p2,
  width = 8,
  height = 5,
  units = "in",
  dpi = 300
)

ggsave(
  filename = "output_plots/quarterly_revenue_institution_line.jpeg",
  plot = p3,
  width = 8,
  height = 5,
  units = "in",
  dpi = 300
)

```

## Summary Checklist
* [x] Data Wrangling: Filtered out test accounts & unpaid transactions using dplyr::filter().

* [x] Piping: Chained operators cleanly using the native R pipe (|>).

* [x] Date Parsing: Extracted month and quarter attributes using lubridate.

* [x] Monthly Trends: Visualized monthly dips via bar charts.

* [x] Quarterly Analysis: Segmented revenue by institution type using stacked bar and line charts.

* [x] Export: Configured publication-quality exports using ggsave().
